---
title: "quiz3"
author: "HH"
date: "11/06/2014"
output: html_document
---
Q1: μ=0.55, σ=1

Q2: In order to reduce the standard error by half, sample size should be doubled.

Q3:the variability of sample statistics.

Q4: Sampling distributions get closer to normality as the sample size increases

Q5:
I false
II true
III true

Q6:
H0:μ=1,800
HA:μ>1,800

Q7:
P(observed or more extreme sample statistic | H0 true)

Q8: (1.6,4.4)
````{r}
abs(-0.5 - 0.9)/2
abs(20.3 - 21.7)/2
abs(1.6 - 4.4)/2
abs(-4.7 - -3.3)/2
````

Q9: < or >

Q10: not rejected when it is false.

Q11: (I), because the sample may not be random and hence observations may not be independent

I.- problem.
II.- ok
III.- ok
IV.- small problem.

Q12: 554

${ME=z^*\dfrac{s}{\sqrt{n} } \rightarrow }$
${n = \dfrac{z^*s}{ME} }$

````{r}
conf=.95;
s = 300;
me=25;

zstar = abs(qnorm((1 - conf)/2));
n=((zstar * s)/me)^2;
sprintf("zstar=%.4f, n=%.4f",zstar, n);
````

Q13: false

Q14: The p-value is approximately 16%, which means we should fail to reject the null hypothesis and determine that these data do not provide convincing evidence the nutrition label does not provide an accurate measure of calories in the bags of potato chips.

````{r}
## Population
mu = 130 # population mean on null hypothesis
sd=17;
alpha=.10;

# Sample
n = 35;
mean = 134; # Sample mean.
SE= sd/sqrt(n); # Standard Error of the sample
zscore=(mean - mu)/SE;

# pvalue = 1 - pnorm(zscore); # One sided.
pvalue = (1 - pnorm(zscore)) * 2; # Two sided.

sprintf("SE=%.4f, zscore=%.4f, p-value=%.4f, reject H0=%s", SE, zscore, pvalue,  pvalue < alpha);

````

Q15:

````{r}
conf=.95;
s = 220;

zstar = 1.96;
n=100;
mean=415;
se=s/sqrt(n);
me=zstar*se;

  
sprintf("se=.4%f, me=%.4f, (%.4f, %.4f)  ",zstar, n, mean - me, mean + me);
````


