---
title: "Quiz4"
output: html_document
---

Q1: (3.75mm, 15mm) Counted dots left and right.

Q2: Assess gender-related salary gap by comparing salaries of randomly sampled men and women.

Q3: sampling distribution

Q4: 

We are 95% confident that overweight children on average miss 1.3 to 2.8 days more than children with normal weight.

Q5: 
A confidence interval based on this sample is not accurate since the sample size is less than 30.

Q6: It becomes more normal looking.


Q7: Tom has a one-sided alternative hypothesis and should do a paired t-test.

Q8: greater than 0.2

````{r}
n = 18;
mu = 3;
t = .5;
pvalue = (1 - pt(t, df = n - 1)); # Two tails.

sprintf("t=%.4f, pvalue=%.4f, rejectH0=%s", t, pvalue, (pvalue < 0.05));
````

Q9: F-test (ANOVA)

Q10: Side-by-side box plots showing roughly equally sized boxes for each group.

Q11: 1.87

````{r}
Fvalue = 75/40.13;
pvalue = pf(Fvalue, 7, 189, lower.tail = FALSE);
sprintf("Group: DFG=%.4f, SSG=%.4f, MSG=%.4f, Fvalue=%.4f,pvalue=%.4f", 7, 525, 75, Fvalue,pvalue);

````

Q12: Ho u1=u2=u3=u4, HA: At least one u is different.

Q13: 1.023
````{r}
x1 = 10;
s1 = 3.4;
n1 = 18;

x2 = 11;
s2 = 2.7;
n2 = 18;

df = min(n1 - 1, n2 - 1);
se = sqrt( s1^2/n1 + s2^2/n2 );
se
````

Q14: 10

````{r}
k=5;
K=k*(k-1)/2;
K
````
