---
title: "unit3"
author: "HH"
date: "11/06/2014"
output: html_document
---

# Unit 3: Introduction.

## Unit 3: Part 1: (1) Sampling Variability and CLT

## Unit 3: Part 1: (2) CLT (for the mean) examples\

## Unit 3: Part 2: (1) Confidence Interval (for a mean)

A confidence interval is a plausible range of values for the population parameter.

### Central Limit Theorem(CLT):

${\bar{x} \tilde{} N(mean = \mu, SE = \dfrac{\sigma}{\sqrt{n}} )}$

### Confidence interval for a population mean:

Computed as the sample mean plus/minus a margin of error (critical value corresponding to the middle XX% of the normal distribution times the standard error of the sampling distribution).

${ \bar{x} \pm z^* \dfrac{s}{\sqrt{n}}  }$

#### Conditions

  1. Independence
    
    * random sample/assigment
    
    * if sample without replacement, n < 10% of population.
    
  2. Sample size/skew
   
    * n >= 30, larger if the population distribution is very skewed.
    
````{r}
conf=.98;
zstar = abs(qnorm((1 - conf)/2));
sprintf("zstar=%.4f", zstar);

sd=20;
n=10;
zstar*(sd/sqrt(n));


````

## Unit 3, Part 2: (2) Accuracy vs precision

### Confidence level

When the confidence level increases the accuracy increases but the precision decreases.

## Unit 3, Part 2: (3) Required sample size for ME

${ME=z^*\dfrac{s}{\sqrt{n} } \rightarrow }$
${n = \dfrac{z^*s}{ME} }$

````{r}
conf=.90;
s = 18;
me=4;

zstar = abs(qnorm((1 - conf)/2));
n=((zstar * s)/me)^2;
sprintf("zstar=%.4f, n=%.4f ",zstar, n);
````
## Unit 3, Part 2: (4) CI (for the mean) examples

## Unit 3, Part 3: (1) Another Introduction for Inference

1. Nothing is going on. Due to chance. Null hypothesis.

2. There is something going on. e.g Discrimination. Alternative hypothesis.

### HT framework

- Start with a null hypothesis (${H_0}) that represents the status quo.

- We also have an alternate hypothesis that represents our research question.

- We conduct a hypothesis test under the assumption that the null hypothesis is true.

- If the test results suggest that the data do not provide convincing evidence for the alternative hypothesis, we stick with the null hypothesis. If we do, then we reject the null hypothesis in favor of the alternative.

## Unit 3, Part 3: (2) Hypothesis Testing (for a mean)

 ${H_0}$ Claim to be tested.
 
 ${H_A}$ 
 
 The skeptic will not abandon the ${H_0}$ unless the evidence in favor of the ${H_A}$ is so strong that she rejects ${H_0}$ in favor of ${H_A}$
 
## Unit 3, Part 3: (3) HT (for the mean) examples

${H_0 \mu  = 3}$

${H_A \mu > 3}$

p-value is  P(observed or more extreme outcome | ${H_0}$ true)

````{r}
## Population
mu = 3; # population mean on null hypothesis
sd=1.74; # Population standard deviation.
alpha=.10;

# Sample
n = 50;
mean = 3.2; # Sample mean.
SE= s/sqrt(n); # Standard Error of the sample
zscore=(mean - mu)/SE;

pvalue = 1 - pnorm(zscore); # One sided.
# pvalue = (1 - pnorm(zscore)) * 2; # Two sided.

sprintf("SE=%.4f, zscore=%.4f, p-value=%.4f, reject H0=%s", SE, zscore, pvalue,  pvalue < alpha);

````

### Decision based on the p-value.

 - if the p-value is very low (lower than ${\alpha}$) we reject ${H_0}$

 - otherwise. Fail to reject ${H_0}$  

## Unit 3, Part 4: Inference for Other Estimators.

## Unit 3, Part 5: (1) Decision Errors.






