---
title: "Unit5"
output: html_document
---

## Unit 5 

#### Inference for Categorical Variables.

Parameter of interest is a proportion.

### Part 1

#### Sampling Variability and the CLT for Proportions

##### CLT for proportions

The distribution of sample proportions is nearly normal, centered at the population proportion, and with a standard error inversely proportional to the sample size.

${SE = \sqrt{\dfrac{p(1-p)}{n}} }$

${ mean = p }$

${\hat{p} \tilde{} N (mean, SE) }$

##### Conditions for the CLT.

1. Independence.

2. Sample size/skew: There should be at least 10 successes and 10 failures in the sample: ${ np \geq 10 }$ and ${ n(1-p) \geq 10 }$

If p is unknown, use ${ \hat{p} }$

````{r}
p = 0.90;
n = 200;

if (n * p < 10 || n * (1-p) < 10)
  "Cannot use this method";

mean=p;
SE=sqrt(p*(1-p)/n);
Z=(0.95 - mean)/SE;
pvalue=1-pnorm(Z);
# pvalue=1-pnorm(0.95, mean, SE); Alternative.

sprintf("SE=%.4f, Z=%.4f, pvalue=%.8f",SE, Z, pvalue);

````
#### Confidence Interval for a Proportion

##### Estimating a proportion

point estimate ${\pm}$ margin of error

${ \hat{p} \pm z\star SE_{\hat{p}} }$

##### Standard Error for a proportion

${SE_{\hat{p}} = \sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}} }$

````{r}
p = 0.85;
n = 670;
SE=sqrt(p*(1-p)/n);
mean=p;
me = 1.96*SE;
sprintf("mean=%.4f, SE=%.4f, (%.4f, %.4f)",p,SE,mean - me, mean + me);

````

##### Calculating the required sample size for desired ME.

${ ME = z^\star\sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}} }$

If there is a previous study that we can rely on for the value of ${ \hat{p} }$ use that in the calculation of the required sample size.

If not, use ${ \hat{p} = 0.5 }$

#### Hypothesis Test for a Proportion.

1. Set the hypothesis:

  - ${H_0: p = null value }$
  
  - ${H_A: p < }$ or ${ > }$ or ${ \ne }$ null value 
  
2. Check conditions:

  - Independence.
  
  - Sample size/skew. ${ np \geq 10 }$ and ${ n(1-p) \geq 10 }$

3. Draw sampling distribution, shade p-value, calculate test statistic.

${ Z = \dfrac{\hat{p} - p}{SE} }$, ${SE = \sqrt{\dfrac{p(1-p)}{n}} }$

4. Make a decision, and interpret it in context of the research question:

- If p-value < ${ \alpha }$, reject ${H_0}$; The data provide convincing evidence for ${H_A}$.\

- If p-value > ${ \alpha }$, fail to reject ${H_0}$; the data do not provide convincing evidence for ${H_A}$.

##### When do we use ${\hat{p} }$ vs ${p}$

- Use ${\hat{p}}$ when you don't have ${p}$.

  * Confidence intervals. Use ${\hat{p}}$.
  
  * Hypothesis test. Use ${p}$.

````{r}
p = 0.5;
phat = 0.6;
n = 1983
alpha=0.05;

SE=sqrt(p*(1-p)/n);
Z= (phat - p)/SE;

pvalue=1 - pnorm(Z);
sprintf("SE=%.4f, Z=%.4f, pvalue=%.4f, rejectH0=%s", SE, Z, pvalue, (pvalue < alpha));

````
### Part 2

#### Estimating the Difference Between Two Proportions.

point estimate ${\pm}$ margin of error

${(\hat{p}_1 - \hat{p}_2 ) \pm z\star SE_{(\hat{p}_1 - \hat{p}_2 )} }$

${ SE = \sqrt{ \dfrac{\hat{p}_1(1 - \hat{p}_1)}{n_1} + \dfrac{\hat{p}_2(1 - \hat{p}_2)}{n_2}} }$

##### Conditions for inference

1. Independent

  * Within gropups.
  
  * Between groups.
  
2. Sample skew. 

  * ${ n_1p_1 \geq 10 }$ and ${ n_1(1-p_1) \geq 10 }$ 
  
  * ${ n_2p_2 \geq 10 }$ and ${ n_2(1-p_2) \geq 10 }$

````{r}
p1 = .71;
n1 = 83;
p2 = .25
n2 = 1028;

pe = p1 - p2; # point estimate.
SE = sqrt(p1*(1-p1)/n1 + p2*(1-p2)/n2);
me = 1.96 * SE;

sprintf("pe=%.4f, SE=%.4f, (%.4f, %.4f)", pe, SE, pe - me, pe + me);



````

#### Hypothesis Test for Comparing Two Proportions.

##### Pooled proportion

${ \hat{p_{pool}} = \dfrac{\text{# of successes}_1 + \text{# of successes}_2}{n_1 + n_2} }$

##### Success-failure condition.

${n_1\hat{p_{pool}} \geq 10}$

${n_1(1 - \hat{p_{pool}}) \geq 10}$


${n_2\hat{p_{pool}} \geq 10}$

${n_2(1 - \hat{p_{pool}}) \geq 10}$

##### Standard Error

${ SE = \sqrt{ \dfrac{\hat{p}_{pool}(1 - \hat{p}_{pool})}{n_1} + \dfrac{\hat{p}_{pool}(1 - \hat{p}_{pool})}{n_2}} }$

````{r}
s1 = 34;
s2 = 61;
p1=0.38;
p2=0.50;
n1 = 90;
n2 = 122;

pool = (s1 + s2)/(n1 + n2);

if (n1*pool < 10)  "success-failure condition failed";
if (n1*(1 - pool) < 10)  "success-failure condition failed";
if (n2*pool < 10) "success-failure condition failed";
if (n2*(1 - pool) < 10)  "success-failure condition failed";

pe = p1 - p2; # point estimate.
SE = sqrt(pool*(1-pool)/n1 + pool*(1-pool)/n2);
z = (pe - 0)/SE;
pvalue=(1 - pnorm(abs(z))) * 2;
sprintf("pe=%.4f, SE=%.4f, z=%.4f, pvalue=%.4f, rejectH0=%s", pe, SE, z, pvalue, pvalue < 0.05);
````

### Part 3

#### Small Sample Proportion

  * What if success-failure condition is not met?
  
  * Inference via simulation.
  
  * Setting up a simultation assuming ${H_0}$ is true.
  
##### Paul the Octopus example

###### Set up hypothesis

${H_0 : p = 0.5}$

${H_A: p > 0.5 }$

${ n = 8 }$

${ \hat{p} = 1 }$

###### Check conditions.

  1. Independence. (Met)

  2. Sample size/skew. (Not met)
  
###### Use simulations.
 
* Devise a simulation scheme that assumes the null hypothesis is true.

* Repeat the simulation many times and record relevant sample statistic.

* Calculate p-value as the proportion of simulations that yield a result favorable to the alternative hypothesis.

````{r}
source("http://bit.ly/dasi_inference")
paul = factor(c(rep("yes", 8), rep("no",0)), levels = c("yes", "no"))
inference(paul, est = "proportion", type="ht", method="simulation", success="yes", null = 0.5, alternative ="greater")
````
